Interspecies comparisons of Mg/Ca ratios in limpet shells

### 2024-06-23 Manuscript.qmd
added modern isotope data

D18O <- read_csv(here("data","D18O.csv"))
MgCa <- read_csv(here("data","MgCa.csv"))

@@

On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009. Westray lies about 70 km north of the Scottish mainland and experiences virtually no freshwater input with salinity values of 34.0 – 34.5 psu. The annual SST ranges from `r WestrayMin`ºC in March to `r WestrayMax`ºC in August (@fig-SSTs).

```{r SST plot}
#| fig-cap: SST values averaged from  years. SST values were sourced from [@europeanunion-copernicusmarineservice2015]
#| fig-height: 10
#| fig-width: 10
#| label: fig-SSTs

@@

@@

>>>>

D18O <- read_csv(here("data","D18O.csv")) %>% 
  mutate(period=case_when(startsWith(sampleID,"Q") ~ "Archaeological",TRUE ~ "Modern"))

MgCa <- read_csv(here("data","MgCa.csv"))

@@

On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009. Westray lies about 70 km north of the Scottish mainland and experiences virtually no freshwater input with salinity values of 34.0 – 34.5 psu [@Inall2009-ho]. The annual SST ranges from `r WestrayMin`ºC in March to `r WestrayMax`ºC in August (@fig-SSTs).

```{r SST plot}
#| fig-cap: SST values averaged from  years. SST values were sourced from [@europeanunion-copernicusmarineservice2015]
#| fig-height: 5
#| fig-width: 5
#| label: fig-SSTs

@@

All shells show multi-year quasi-sinusoidal records of SST change (@fig-Mod_iso), with *Nacella* sp. shells living 2–3 years and *Patella v*. over 5 years. The δ^18^O values of the Orcadian shell have not been directly compared to the local SST, but accurately reflect the seasonal changes expected with one large anomaly in the second summer of recorded growth. <!--# ADD FIGURE TO THESE ISOTOPE VALUES -->

```{r Modern Isotope data}
#| fig-cap: Isotope values of modern limpet shells from Cambaceres Bay (*a-f*) and Westray (*g*).
#| fig-height: 10
#| fig-width: 10
#| label: fig-Mod_iso
# Filter data for Modern period
modern_data <- D18O %>% filter(period == "Modern")

# Get unique sample IDs
sample_ids <- unique(modern_data$sampleID)

# Initialise an empty list to store plots
plots <- list()

# Loop over each sampleID to create individual plots
for (i in seq_along(sample_ids)) {
  sample_id <- sample_ids[i]
  plot <- modern_data %>%
    filter(sampleID == sample_id) %>%
    ggplot(aes(dist, d18O)) +
    geom_path() +
    geom_point() +
    xlim(0,15)+
    labs(
      x = "Distance from margin in mm",
      y = expression(delta^18*O~("\u2030")),
      subtitle = sample_id,
      col = "",
      fill = ""
    ) +
    theme(plot.subtitle = element_text(hjust = 0.5))

  # Store plot in the list
  plots[[i]] <- plot
}

# Combine plots using patchwork and add letters
wrap_plots(plots,nrow=4) + plot_annotation(tag_levels = 'a')

@@

In total, we studied 16 archaeological specimens. Of those, seven come from Quoygrew on Westray dating to 900 – 1200 CE. These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly and the Little Ice Age [@Surge2012-ba]. and nine from the site Hesk

### 2024-06-23 Manuscript.tex
added modern isotope data

There are two main locations that we sourced specimens from: the Beagle
Channel in Tierra del Fuego (Argentina) and the island of West Ray in
Orkney (UK). Four specimens were collected from the Beagle Channel in
Cambaceres Bay. The area receives around 570 mm precipitation but the
channel waters are predominantly influenced by marine currents stayed
around 30.7 ± 0.7 psu \citep{Nicastro2020-ih}. SSTs range from 4.7ºC in
August to 10ºC in February.

On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in
August 2009. Westray lies about 70 km north of the Scottish mainland and
experiences virtually no freshwater input. The annual SST ranges from
6.3ºC in March to 13.8ºC in August (Figure~\ref{fig-SSTs}).

@@

@@

>>>>

There are two main locations that we sourced specimens from: the Beagle
Channel in Tierra del Fuego (Argentina) and the island of West Ray in
Orkney (UK). Four specimens were collected from the Beagle Channel in
Cambaceres Bay. The area receives around 570 mm precipitation but the
channel waters are predominantly influenced by marine currents and in
the period of October 2015 to October 2016 stayed around 30.7 ± 0.7 psu
\citep{Nicastro2020-ih}. SSTs range from 4.7ºC in August to 10ºC in
February.

On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in
August 2009. Westray lies about 70 km north of the Scottish mainland and
experiences virtually no freshwater input with salinity values of 34.0
-- 34.5 psu \citep{Inall2009-ho}. The annual SST ranges from 6.3ºC in
March to 13.8ºC in August (Figure~\ref{fig-SSTs}).

@@

All shells show multi-year quasi-sinusoidal records of SST change
(Figure~\ref{fig-Mod_iso}), with \emph{Nacella} sp. shells living 2--3
years and \emph{Patella v}. over 5 years. The δ\textsuperscript{18}O
values of the Orcadian shell have not been directly compared to the
local SST, but accurately reflect the seasonal changes expected with one
large anomaly in the second summer of recorded growth.

\includegraphics{Manuscript_files/figure-pdf/fig-Mod_iso-1.pdf}

\caption{\label{fig-Mod_iso}Isotope values of modern limpet shells from
Cambaceres Bay (\emph{a-f}) and Westray (\emph{g}).}

@@

In total, we studied 16 archaeological specimens. Of those, seven come
from Quoygrew on Westray dating to 900 -- 1200 CE. These shells have
previously been studied to characterise seasonal temperature change
during the Medieval Climate Anomaly and the Little Ice Age
\citep{Surge2012-ba}. and nine from the site Hesk
